Open source landscape ecology tools \thanks{Grants or other notes about the article that should go on the front page should be placed here. General acknowledgments should be placed at the end of the article.}

### 2020-07-14 paper/paper.Rmd
Some more progress on existing packages

---
title: Open source landscape ecology tools
# subtitle: Do you have a subtitle? If so, write it here
# titlerunning: Open source landscape ecology tools
# authorrunning: Short form of author list if too long for running head

thanks: | 
    Grants or other notes about the article that should go on the front 
    page should be placed here. General acknowledgments should be placed at the
    end of the article.

authors: 
- name: Maximillian H.K. Hesselbarth
  address: Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
  email: maximilian.hesselbarth@uni-goettingen.de
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Author 3
  address: Adress author 3
  email: E-Mail author 3
- name: ...
  address: ...
  email: ...

keywords:
- keyword 1
- keyword 2
- keyword 3
- keyword 4
- ...

abstract: |
  max. (200 words) Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract

bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
output: rticles::springer_article
---

@@

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}. Open-source software includes all software which is released under a license that allows to freely use, modify and distribute the software \cite{St.Laurent2008}. Open-source software development has many advantages, such as fast innovation, transparency and reliability as well as longevity mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}. Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology \cite{Prlic2012,Powers2019}. Lastly, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} and customize existing software to their specific needs easily \cite{Steiniger2009}.

@@

Species distribution modeling (SDM) analysis how landscape patterns (e.g. habitat suitability, resources availability, etc.) influence and determine the patterns of species distribution, mainly to infer ecological processes and predict future species distributions \cite{Wiersma2011}. Originated in the 1970s, SDM has experienced numerous methodological advancement and a numerous body of literature exists today \cite{Zimmermann2010}. Additionally, textbooks introducing  basic concepts of SDM in R exists \cite{Guisan2017,Fletcher2019}.

Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of *R* packages used for SDMs. Popular approaches and packages include generalized linear models using e.g. the *stats* package; generalized additive models using e.g.the *mgcv* or *lme4* package; classification and regression trees (CART) using e.g. the *rpart*, *randomForest* or *ranger* package or multivariate data analysis using e.g. the *ade4* package. Another widely used SDM approach uses the concept of maximum entropy \cite{Phillips2006,Fletcher2019}. However, currently the *maxent* package is not available on *CRAN* anymore. Of course, also packages specifically designed for SDM exists. This includes the *dismo*, *sdm*, *ecospat*, *biomod2* and *PresenceAbsence* packages.

@@

Given its wide concept, many different measures of connectivity exists \cite{Kindlmann2008}. On patch level structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g. amount of suitable habitat) \cite{Kindlmann2008,With2019}. Such measures are provided within the *landscapemetrics* package (see \ref{sec:landscape_metrics}). Another way to describe connectivity is based on graph theory with the advantages that also functional connectivity can be included \cite{Kindlmann2008}. In graph theory  \cite{Laita2011} landscapes are described by nodes (i.e. habitat patches) connected by and functional connections called links \cite{Laita2011}. The *grainscape* package \cite{Chubaty2020} provides a tool to model connectivity based on spatially explicit networks. More general, the *igraph* packages \cite{Csardi2006} provides functionality related to graph theory.

<!-- MH: Work-in-Progress -->

<!-- - resistance surfaces (DOI: https://doi.org/10.1016/S0169-2046(02)00242-6): *gdistance* package DOI: 10.18637/jss.v076.i13-->

<!-- - *samc*: an R package for connectivity modeling with spatial absorbing Markov chains; DOI: https://doi.org/10.1111/ecog.04891 -->

@@

<!-- MH: Work-in-Progress -->
<!-- There seems to be a lot of connection to basic connectivity -->
<!-- ISOLATION BY RESISTANCE; DOI: 10.1111/j.0014-3820.2006.tb00500.x -->
<!-- *graph4lg*  -->
<!-- *gdistance* -->
<!-- *lme4* -->
<!-- *HierDpart* -->
<!-- *PopGenReport* -->
<!-- https://dyerlab.github.io/Landscape-Genetics-Data-Analysis/index.html -->

#### Neutral landscape models {#sec:NLM}

<!-- MH: NLMR; DOI:  https://doi.org/10.1111/2041-210X.13076 -->

>>>>

---
title: Open source landscape ecology tools
# subtitle: Do you have a subtitle? If so, write it here
# titlerunning: Open source landscape ecology tools
# authorrunning: Short form of author list if too long for running head
thanks: | 
    Grants or other notes about the article that should go on the front 
    page should be placed here. General acknowledgments should be placed at the
    end of the article.
authors: 
- name: Maximillian H.K. Hesselbarth
  address: Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
  email: maximilian.hesselbarth@uni-goettingen.de
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Author 3
  address: Adress author 3
  email: E-Mail author 3
- name: ...
  address: ...
  email: ...
keywords:
- keyword 1
- keyword 2
- keyword 3
- keyword 4
- ...
abstract: |
  max. (200 words) Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
output: rticles::springer_article
editor_options: 
  chunk_output_type: console
---

@@

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}. Open-source software includes all software which is released under a license that allows to freely use, modify and distribute the software \cite{St.Laurent2008}. Open-source software development has many advantages, such as fast innovation, transparency and reliability as well as longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}. Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology \cite{Prlic2012,Powers2019}. Lastly, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} and to customize existing software to their specific needs \cite{Steiniger2009}.

@@

Species distribution modeling (SDM) analysis how landscape patterns (e.g. habitat suitability, resources availability, etc.) influence and determine the patterns of species distribution, mainly to infer ecological processes and predict future species distributions \cite{Wiersma2011}. Originated in the 1970s, SDM has experienced numerous methodological advancement and a numerous body of literature exists today \cite{Zimmermann2010}. Additionally, textbooks introducing  basic concepts of SDM in *R* exists \cite{Guisan2017,Fletcher2019}.

Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of *R* packages used for SDMs. Popular approaches and packages include generalized linear models using e.g. the *stats* package; generalized additive models using e.g.the *mgcv* or *lme4* package; classification and regression trees (CART) using e.g. the *rpart*, *randomForest* or *ranger* package or multivariate data analysis using e.g. the *ade4* or *vegan* package. Of course, also packages specifically designed for SDM exists. This includes the *dismo*, *sdm*, *ecospat*, *biomod2* and *PresenceAbsence* packages.

@@

Given its wide concept, many different measures of connectivity exists \cite{Kindlmann2008}. On patch level structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g. amount of suitable habitat) \cite{Kindlmann2008,With2019}. Such measures are provided within the *landscapemetrics* package (see \ref{sec:landscape_metrics}). Another way to describe connectivity is based on graph theory with the advantages that also functional connectivity can be included \cite{Kindlmann2008}. In graph theory  \cite{Laita2011} landscapes are described by nodes (i.e. habitat patches) connected by and functional connections called links \cite{Laita2011}. The *grainscape* package \cite{Chubaty2020} provides a tool to model connectivity based on spatially explicit networks. More general, the *igraph* packages \cite{Csardi2006} provides functionality related to graph theory. Resistance surfaces and least-cost paths are another tool to model functional connectivity. The resistance surface describes effects of facilitating or hindering characteristics of the landscape for an organism moving within it \cite{Adriaensen2003}. Least-cost paths can be calculated using the *gdistance* package \cite{vanEtten2017}. Absorbing Markov chains that combines random walk theory and mortality based on the landscape characteristics \cite{Fletcher2019a} and is provided by the recently published *samc* package \cite{Marx2020}.

@@

<!-- MH: Work-in-Progress -->
<!-- MH: General population genetics -->
<!-- MH: Isolation by resistance  -->
<!-- MH: Bayesian Genetic Clustering -->
<!-- MH: Multivariate Ordination Techniques -->

Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:landscape_genetics}) are also of importance for landscape genetics. Further functionality for landscape genetics can be found in the *graphs4lg* \cite{Savary2020}, *PopGenReport* \cite{Adamack2014,Gruber2015}, or *HierDpart* packages \cite{Qin2019}.

#### Neutral landscape models {#sec:NLM}
Neutral landscape models are used to create structured landscapes in the absence of specific ecological and landscape processes as null models against which hypotheses including specific ecological and landscape processes can be tested statistically \cite{Gardner1987,With1997}. Because neutral landscape models are not based on ecological and landscape process, many different generic algorithms to create landscapes can be found (for a comprehensive list, see \cite{Sciaini2018}).

Currently, to our knowledge, only one *R* packages exists that allows the use of neutral landscapes. Using a wide collection of algorithms, the *NLMR* package \cite{Sciaini2018} provides framework to simulate neutral landscape rasters.

<!-- # Missing packages -->

<!-- Another widely used SDM approach uses the concept of maximum entropy \cite{Phillips2006,Fletcher2019}. However, currently the *maxent* package is not available on *CRAN* anymore. -->

### 2020-07-14 paper/paper.tex
Some more progress on existing packages

Software to manage and analyze data becomes increasingly important in
modern scientific research \cite{Wilson2014} and many scientific studies
would not have been possible without open-source software
\cite{Prlic2012}. Open-source software includes all software which is
released under a license that allows to freely use, modify and
distribute the software \cite{St.Laurent2008}. Open-source software
development has many advantages, such as fast innovation, transparency
and reliability as well as longevity mainly due to many diverse
contributors \cite{vonKrogh2006,St.Laurent2008}. Additionally, the use
of open-source software facilitates (computational) reproducibility and
can allow a better understanding of the methodology
\cite{Prlic2012,Powers2019}. Lastly, open-source software allows other
scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012}
and customize existing software to their specific needs easily
\cite{Steiniger2009}.

@@

Species distribution modeling (SDM) analysis how landscape patterns
(e.g.~habitat suitability, resources availability, etc.) influence and
determine the patterns of species distribution, mainly to infer
ecological processes and predict future species distributions
\cite{Wiersma2011}. Originated in the 1970s, SDM has experienced
numerous methodological advancement and a numerous body of literature
exists today \cite{Zimmermann2010}. Additionally, textbooks introducing
basic concepts of SDM in R exists \cite{Guisan2017,Fletcher2019}.

Because the used modeling approaches are diverse
\cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of
\emph{R} packages used for SDMs. Popular approaches and packages include
generalized linear models using e.g.~the \emph{stats} package;
generalized additive models using e.g.the \emph{mgcv} or \emph{lme4}
package; classification and regression trees (CART) using e.g.~the
\emph{rpart}, \emph{randomForest} or \emph{ranger} package or
multivariate data analysis using e.g.~the \emph{ade4} package. Another
widely used SDM approach uses the concept of maximum entropy
\cite{Phillips2006,Fletcher2019}. However, currently the \emph{maxent}
package is not available on \emph{CRAN} anymore. Of course, also
packages specifically designed for SDM exists. This includes the
\emph{dismo}, \emph{sdm}, \emph{ecospat}, \emph{biomod2} and
\emph{PresenceAbsence} packages.

@@

Given its wide concept, many different measures of connectivity exists
\cite{Kindlmann2008}. On patch level structural connectivity can be
measured using nearest-neighbor distances or characterizations of the
patch neighborhood (e.g.~amount of suitable habitat)
\cite{Kindlmann2008,With2019}. Such measures are provided within the
\emph{landscapemetrics} package (see \ref{sec:landscape_metrics}).
Another way to describe connectivity is based on graph theory with the
advantages that also functional connectivity can be included
\cite{Kindlmann2008}. In graph theory \cite{Laita2011} landscapes are
described by nodes (i.e.~habitat patches) connected by and functional
connections called links \cite{Laita2011}. The \emph{grainscape} package
\cite{Chubaty2020} provides a tool to model connectivity based on
spatially explicit networks. More general, the \emph{igraph} packages
\cite{Csardi2006} provides functionality related to graph theory.

@@

@@

>>>>

Software to manage and analyze data becomes increasingly important in
modern scientific research \cite{Wilson2014} and many scientific studies
would not have been possible without open-source software
\cite{Prlic2012}. Open-source software includes all software which is
released under a license that allows to freely use, modify and
distribute the software \cite{St.Laurent2008}. Open-source software
development has many advantages, such as fast innovation, transparency
and reliability as well as longevity, mainly due to many diverse
contributors \cite{vonKrogh2006,St.Laurent2008}. Additionally, the use
of open-source software facilitates (computational) reproducibility and
can allow a better understanding of the methodology
\cite{Prlic2012,Powers2019}. Lastly, open-source software allows other
scientists to reuse code and not ``reinvent the wheel'' \cite{Prlic2012}
and to customize existing software to their specific needs
\cite{Steiniger2009}.

@@

Species distribution modeling (SDM) analysis how landscape patterns
(e.g.~habitat suitability, resources availability, etc.) influence and
determine the patterns of species distribution, mainly to infer
ecological processes and predict future species distributions
\cite{Wiersma2011}. Originated in the 1970s, SDM has experienced
numerous methodological advancement and a numerous body of literature
exists today \cite{Zimmermann2010}. Additionally, textbooks introducing
basic concepts of SDM in \emph{R} exists \cite{Guisan2017,Fletcher2019}.

Because the used modeling approaches are diverse
\cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of
\emph{R} packages used for SDMs. Popular approaches and packages include
generalized linear models using e.g.~the \emph{stats} package;
generalized additive models using e.g.the \emph{mgcv} or \emph{lme4}
package; classification and regression trees (CART) using e.g.~the
\emph{rpart}, \emph{randomForest} or \emph{ranger} package or
multivariate data analysis using e.g.~the \emph{ade4} or \emph{vegan}
package. Of course, also packages specifically designed for SDM exists.
This includes the \emph{dismo}, \emph{sdm}, \emph{ecospat},
\emph{biomod2} and \emph{PresenceAbsence} packages.

@@

Given its wide concept, many different measures of connectivity exists
\cite{Kindlmann2008}. On patch level structural connectivity can be
measured using nearest-neighbor distances or characterizations of the
patch neighborhood (e.g.~amount of suitable habitat)
\cite{Kindlmann2008,With2019}. Such measures are provided within the
\emph{landscapemetrics} package (see \ref{sec:landscape_metrics}).
Another way to describe connectivity is based on graph theory with the
advantages that also functional connectivity can be included
\cite{Kindlmann2008}. In graph theory \cite{Laita2011} landscapes are
described by nodes (i.e.~habitat patches) connected by and functional
connections called links \cite{Laita2011}. The \emph{grainscape} package
\cite{Chubaty2020} provides a tool to model connectivity based on
spatially explicit networks. More general, the \emph{igraph} packages
\cite{Csardi2006} provides functionality related to graph theory.
Resistance surfaces and least-cost paths are another tool to model
functional connectivity. The resistance surface describes effects of
facilitating or hindering characteristics of the landscape for an
organism moving within it \cite{Adriaensen2003}. Least-cost paths can be
calculated using the \emph{gdistance} package \cite{vanEtten2017}.
Absorbing Markov chains that combines random walk theory and mortality
based on the landscape characteristics \cite{Fletcher2019a} and is
provided by the recently published \emph{samc} package \cite{Marx2020}.

@@

Since describing connectivity between two locations is one of the
fundamental steps of landscape genetics, all packages useful for
connectivity (see \ref{sec:landscape_genetics}) are also of importance
for landscape genetics. Further functionality for landscape genetics can
be found in the \emph{graphs4lg} \cite{Savary2020}, \emph{PopGenReport}
\cite{Adamack2014,Gruber2015}, or \emph{HierDpart} packages
\cite{Qin2019}.

@@

Neutral landscape models are used to create structured landscapes in the
absence of specific ecological and landscape processes as null models
against which hypotheses including specific ecological and landscape
processes can be tested statistically \cite{Gardner1987,With1997}.
Because neutral landscape models are not based on ecological and
landscape process, many different generic algorithms to create
landscapes can be found (for a comprehensive list, see
\cite{Sciaini2018}).

Currently, to our knowledge, only one \emph{R} packages exists that
allows the use of neutral landscapes. Using a wide collection of
algorithms, the \emph{NLMR} package \cite{Sciaini2018} provides
framework to simulate neutral landscape rasters.
